---
title: "Creating An Efficient Data Analysis Workflow, Part 2"
author: "Miguel Asse"
date: "12/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Problem

We are taking on the role of as an analyst for a book company. The company has provided us more data on some of its 2019 book sales, and it wants us to extract some usable knowledge from it. It launched a new program encouraging customers to buy more books on July 1st, 2019, and we want to know **if this new program was successful at increasing sales and improving review quality**.

## Data Exploration

Let's load our dataset and look at the data quality

```{r}
library(readr)
library(tidyverse)
data <- read_csv("sales2019.csv")

unique(data)
```

How big is the dataset?
```{r}
dim(data)
```
**Note that total purchased column contains N/As**

How many NAs?
```{r}
colSums(is.na(data))
```
**There are 718 total_purchased NAs and 885 user_submitted_review NAs**

What are the column names?
```{r}
column_names <- colnames(data)
column_names
```

## Data Cleaning

As part of our data cleaning process, we are going to:
  1. Remove any rows with an NA value in user user_submitted_review
  1. We'll fill in the total_purchased columns with the average number of books purchased across the entire dataset
  1. Determine if the sentence in user_submitted review is positive or not based on the text

```{r}
clean_reviews <- data %>%
  dplyr::filter(
    !is.na(user_submitted_review)
  )

dim(clean_reviews)
```

As we expect, 885 rows are removed.

```{r}
average_number_of_books_purchased <- clean_reviews %>%
  filter(
    !is.na(total_purchased)
  ) %>%
  pull(total_purchased) %>%
  mean()
average_number_of_books_purchased
```

Let's fill in the NA values for total_purchased with our calculated mean:
```{r}
clean_reviews <- clean_reviews %>%
  mutate(
    imputed_total_purchased
      = if_else(
        is.na(total_purchased),
        average_number_of_books_purchased,
        total_purchased)
  )

head(clean_reviews, 5)
```

We see the 5th row is NA and has our imputed mean of 3.99 in the imputed_total_purchased column.

Let's create a function to determine if a review is positive or negative based on the text. First let's look at the unique words in our review column:
```{r}
unique(clean_reviews$user_submitted_review)
```
Let's create our positive review function:
```{r}
determine_positive_review <- function(review) {
  review_positive = case_when(
    str_detect(review, "Awesome") ~ TRUE,
    str_detect(review, "Never") ~ TRUE,
    str_detect(review, "okay") ~ TRUE,
    str_detect(review, "a lot") ~ TRUE,
    TRUE ~ FALSE # Set any reviews outside of the above to FALSE
  )
}
```

And let's finally classify each review as positive or not:
```{r}
clean_reviews <- clean_reviews %>%
  mutate(
    is_positive_review = determine_positive_review(user_submitted_review)
  )
```

Let's also clean our data column
```{r}
library(lubridate)
clean_reviews$date <- mdy(clean_reviews$date)
```

Let's determine if a sale happened before or after our program started
```{r}
clean_reviews <- clean_reviews %>%
  mutate(
    sale_after_program_start = clean_reviews$date > mdy("07/01/2019")
  )
```
## Answering Our Business Question

Now that our data is clean, we can determine **if this new program was successful at increasing sales and improving review quality**, since we know the date of our program started on July 1, 2019 and we have the date of the sale to determine whether it happened after our program started.

Overall book sales before and after the program started:
```{r}
book_sales_review <- clean_reviews %>%
  group_by(sale_after_program_start) %>%
  summarise(
    books_purchased = sum(imputed_total_purchased))

book_sales_review
```

It doesn't appear sales have increased after the program started. Is there another way to look at the data, perhaps certain titles?

```{r}
book_sales_review_by_title <- clean_reviews %>%
  group_by(sale_after_program_start, title) %>%
  summarise(
    books_purchased = sum(imputed_total_purchased)) %>%
  arrange(title, sale_after_program_start)
book_sales_review_by_title
```

It looks like "R for Dummmies", "Secrets Of R For Advanced Students" all increased. While "R Made Easy" also increased, the number of books is so small I don't think it's significant enough to call out.

Let's also determine if businesses bought more books versus individuals.

```{r}
book_sales_review_by_customer_type <- clean_reviews %>%
  group_by(sale_after_program_start, customer_type) %>%
  summarise(
    books_purchased = sum(imputed_total_purchased)) %>%
  arrange(customer_type, sale_after_program_start)
book_sales_review_by_customer_type
```

It looks like businesses bought more books than individuals after our program started. There could be an opportunity to further market or expand the program for business customers versus individuals.

Additionally, one final question is **Did review scores improve as a result of the program?**

```{r}
positive_reviews_after_program <- clean_reviews %>%
  group_by(sale_after_program_start) %>%
  summarise(
    total_positive_reviews = sum(is_positive_review)) %>%
  arrange(total_positive_reviews, sale_after_program_start)
```

While positive reviews did slightly increase, the overall effect seems minimal and the program likely did not have any effect.