---
title: "Covid 19 Virus Trends"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

As part of dataquest's data analysis in R path, we are analyzing a COVID-19 dataset using R. We'll be doing so to:

1.  Build the skills and tools of the data analysis worklow of R
1.  Evaluate the positivity of each country, specifically focusing on the question:
    **Which countries have had the highest number of positive cases against the number of tests?**
1. Note that our dataset includes cumulative and daily data mixed together, which we will have to account and filter for in our analysis

## Data Exploration

Let's see the size and shape of our data

```{r}
library(readr)
library(tibble)
library(dplyr)
covid_df <- read_csv("covid19.csv")
dim(covid_df)
```

Here are the column names and first few values of our dataset, as well as data type
```{r}
col_names = colnames(covid_df)
col_names
head(covid_df)
glimpse(covid_df)
```

## Data Filtering

Let's filter our dataset to only include country-level rows, where Province_State is "All States"
```{r}
covid_df_all_states <- covid_df %>%
  filter(Province_State == "All States") %>%
  select(-Province_State)
```

Filter to only include daily data
```{r}
covid_df_all_states_daily <- covid_df_all_states %>%
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive)
```

## Data creation

We also need to create new columns to answer our question: **Which countries have had the highest number of positive cases against the number of tests?**

Let's create a new summary table by country and calculate all their baseline covid statistics

```{r}
covid_df_all_states_daily_sum <- covid_df_all_states_daily %>%
  group_by(Country_Region) %>%
  summarize(
    tested = sum(daily_tested),
    positive = sum(daily_positive),
    active = sum(active),
    hospitalized = sum(hospitalizedCurr)) %>%
  arrange(-tested)

glimpse(covid_df_all_states_daily_sum)
```

Get the top 10 of our results
```{r}
covid_top_10 <- head(covid_df_all_states_daily_sum, 10)
```

Extract the data, name each vector with the associated country, and look at the positivity ratio for the top 10 countries
```{r}
countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$tested
positive_cases <- covid_top_10$positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalized

names(tested_cases) <- countries
names(positive_cases) <- countries
names(active_cases) <- countries
names(hospitalized_cases) <- countries

positive_cases / tested_cases

positive_tested_top_3 <- c("United Kingdom" = 0.11, "United States" = 0.10, "Turkey" = 0.08)
```

Create a matrix of the top 3
```{r}
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)

covid_mat <- rbind(united_kingdom, united_states, turkey)
names <- c("Ratio", "tested", "positive", "active", "hospitalized")
colnames(covid_mat) <- names

covid_mat
```

Create a final summary list of everything with the answer to our question
```{r}
question <- "Which countries have had the highest number of positive cases against the number of tests?"
answer <- c("Positive tested cases" = positive_tested_top_3)

datasets <- list(
  original = covid_df,
  allstates = covid_df_all_states,
  daily = covid_df_all_states_daily,
  top10 = covid_top_10
)

matrices <- list(
  covid_mat
)

vectors <- list(
  col_names,
  countries
)

data_structure_list <- list(
  "dataframe" = datasets,
  "matrix" = matrices,
  "vector" = vectors
)

covid_analysis_list <- list(
  "question" = question,
  "answer" = answer,
  "data" = data_structure_list
)

covid_analysis_list[2]
```

## Conclusion

Which countries have had the highest number of positive cases against the number of tests?

As we see from our filtering and brief analysis above, the UK, USA and Turkey have the highest positivity rate (i.e., positive cases against the number of tests).