---
title: "Forest Fire Data Visualization"
author: "Miguel Asse"
date: "12/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Forest fires can create ecological problems and endanger human lives and
property. Understanding when they occur and what causes them is important for
managing them. The data we'll be working with in this guided project is
associated with a scientific research paper on predicting the occurrence of
forest fires in Portugal using modeling techniques. The dataset can be found
[here](https://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/).

For this project, we will simply be performing some exploratory data analysis
using some of the libraries available in the tidyverse universe of R.

## The Data

Our dataset has the following fields, with definitions below:

* X: X-axis spatial coordinate within the Montesinho park map: 1 to 9
* Y: Y-axis spatial coordinate within the Montesinho park map: 2 to 9
* month: Month of the year: 'jan' to 'dec'
* day: Day of the week: 'mon' to 'sun'
* FFMC: Fine Fuel Moisture Code index from the FWI system: 18.7 to 96.20
* DMC: Duff Moisture Code index from the FWI system: 1.1 to 291.3
* DC: Drought Code index from the FWI system: 7.9 to 860.6
* ISI: Initial Spread Index from the FWI system: 0.0 to 56.10
* temp: Temperature in Celsius degrees: 2.2 to 33.30
* RH: Relative humidity in percentage: 15.0 to 100
* wind: Wind speed in km/h: 0.40 to 9.40
* rain: Outside rain in mm/m2 : 0.0 to 6.4
* area: The burned area of the forest (in ha): 0.00 to 1090.84

Each row represents a fire incident, including the month and day of that month, along data about each fire, based on the fields mentioned above.

## Exploratory Data Analysis

```{r}
library(tidyverse)

data <- read_csv("forestfires.csv")

head(data)

unique(data$month)
unique(data$day)
```

Let's change the month and day to be categorical variables:

```{r}
data <- data %>%
  mutate(
    month_category = factor(
      month,
      levels=c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep",
               "oct", "nov", "dec")),
    day_category = factor(
      day,
      levels=c("mon", "tue", "wed", "thu", "fri", "sat", "sun")
    )
  )
```

## Data Visualization Prep

In order to appropriately visualize fire frequency over time, we need to group
our fires dataset by the month and day so we can appropriately answer questions
such as:

* What days do fires occur most frequently?
* What months do fires occur most frequently?

```{r}
monthly_forest_fires <- data %>%
  group_by(month_category) %>%
  summarise(
    number_of_fires = n()
  )

forest_fires_by_day_of_week <- data %>%
  group_by(day_category) %>%
  summarise(
    number_of_fires = n()
  )
```   

## Data Visualization

### Monthly Forest Fires

```{r}
monthly_forest_fires %>%
  ggplot(aes(x=month_category, y=number_of_fires, group=1)) +
  theme(panel.background = element_rect(fill="white")) + 
  geom_line() +
  labs(
    title="Frequency of Forest Fires By Month",
    x="Month",
    y="Number of Fires"
  )
```

**It appears fires are most common in August and September.**

### Forest Fires By Frequency of Day 

```{r}
forest_fires_by_day_of_week %>%
  ggplot(aes(x=day_category, y=number_of_fires)) +
  theme(panel.background = element_rect(fill="white")) + 
  geom_col(fill="blue") +
  labs(
    title="Frequency of Forest Fires By Day",
    x="Day",
    y="Number of Fires"
  )
```

**Additionally, fires are most likely to occur on Friday, Saturday, or Sunday,
AKA the weekend.**

### Further EDA Analysis

Let's examine all the relationships of the other variables over time. To do that
we need to transform our data into long format and graph all the variables
across the month:

```{r}
data_long_format <- data %>%
  pivot_longer(
    cols = c("FFMC", "DMC", "DC", "ISI", "temp", "RH", "wind", "rain"),
    names_to = "column",
    values_to = "value"
  )

data_long_format %>%
  ggplot(aes(x=month_category, y=value)) +
  geom_boxplot() +
  facet_wrap(vars(column), scales="free_y") +
  labs(
    title="Fire Variable Changes Over Month",
    x="Month",
    y="Scale"
  )
```

**It appears DC, DMC, Temperature, and Wind all increase during the most
frequent fire month of August.**

### Examining Forest Fire Severity

We don't have a variable in our data understandy how severe a forest fire was.
However, a reasonable proxy would be to use the **area** variable to estimate
the impact. As a reminder, the area variable is measuring the burned area of the
forest in hectares (abbreviated as ha).

```{r}
data_long_format %>%
  ggplot(aes(x=value, y=area)) +
  geom_point() +
  facet_wrap(vars(column), scales="free_x") +
  labs(
    title="Fire Variable Changes By Impact To Area",
    x="Variables",
    y="Area (Number of Hectares)"
  )
```

Based on in initial look, it appears rain, and the initial spread (ISI) have a
mitigating impact on the total area finally burned, which makes inherent sense.
Additionally, as the FFMC (Fine Fuel Moisture Code) increases, so does the
overall number of hectares burned.

However, there are some outliers in our data, and the "0s" in our dataset in
terms of area skew our data. Let's re-run our visualization without those
outliers.

```{r}
data %>%
  ggplot(aes(x=area, )) +
  geom_histogram()
```

Let's re-run our above visualization, excluding 0s and area values over 150:

```{r}
data_long_format %>%
  filter(area != 0, area <= 150) %>%
  ggplot(aes(x=value, y=area)) +
  geom_point() +
  facet_wrap(vars(column), scales="free_x") +
  labs(
    title="Fire Variable Changes By Impact To Area",
    x="Variables",
    y="Area (Number of Hectares)"
  )

```

Re-examining, other than FFMC, there doesn't appear to be any significant
relationship between area and the other variables, so it appears our assumption
around area as a proxy for forest fire severity is not a great one.